---
title: 'OWID: Drug Overdose Deaths'
author: "IllicitDavid Jackson"
date: "2/4/2021 06:15 PM/EST"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE)
library(ggplot2)
library(scales)
library(plotly)
library(dplyr)
library(readr)
library(tidyr)
library(forcats)
library(skimr) #Julia Silge 
library(pracma) # Various moving averages
```
```{r}
## Packages needed for creating maps
library(maps)
library(mapproj)
library(sf)
library(tmap)
library(tmaptools)
library(leaflet)
# options(scipen=999)
```

# Drug Overdose Deaths: OUr World In Data (http://ourworldindata.org)
* Death rates from drug overdoses in the United States, measured as the number of deaths per 100,000 individuals (of all ages, not age-standardized).

```{r}
rm(list=ls())
overdose_data <- read_csv("./DATA/deaths-drug-overdoses.csv") %>%
  rename("Country" = "Entity") 

overdose_data$Total <- overdose_data$Opioid + overdose_data$Cocaine + overdose_data$Amphetamine + overdose_data$Other_drug

overdose_world <-overdose_data %>%   filter(Country =="World")
overdose_not_world <-overdose_data %>%   filter(Country !="World")
```

# Plot Total Drug Overdose Death by Year

```{r}
p0 <-overdose_world %>% ggplot(aes(x=Year,y=Total)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="Worldwide Total Deaths Do To Drug Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p0)
```

# Deaths by Drug and Year

```{r}
p1 <-overdose_world %>% ggplot(aes(x=Year,y=Opioid)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="Worldwide Total Deaths Opioid Drug Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p1)
```
```{r}
p2 <-overdose_world %>% ggplot(aes(x=Year,y=Cocaine)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="Worldwide Total Deaths  Drug Cocaine Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p2)
```
```{r}
p3 <- overdose_world %>% ggplot(aes(x=Year,y=Amphetamine)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="Worldwide Total Deaths  Amphetamine Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p3)
```
```{r}
p4 <- overdose_world %>% ggplot(aes(x=Year,y=Other_drug)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="Worldwide Total Deaths Other Drugs Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p4)
```


# Part 2: Use Pivot_wider and plot 


```{r}
od_countrydeaths <- overdose_not_world  %>% group_by(Country) %>%
  summarise(across(Opioid:Total,sum))

```
```{r}
# world_tbl <- map_data("world") %>%
#     as_tibble() %>% rename("Country" = "region") 
# 
# 
# world_join <-   right_join(world_tbl,od_countrydeath)
# world_join <- world_join %>% select(-group,-order,-subregion)
```
```{r}
od_longer <- pivot_longer(overdose_world,
                         cols=c("Opioid","Cocaine","Amphetamine","Other_drug","Total"))
```

### Plot of Drug Total by Year 

```{r}
od_longer %>% filter(name !="Total") %>% ggplot() + geom_line(aes(x=Year,y=value,col=name)) + scale_y_log10(labels=comma) +
  labs(title="Worldwide Overdoes Deaths by Drug and Year",y="Deaths Do TO Overdose") + guides(color = guide_legend(override.aes = list(size = 2)))
```

```{r}
od_longer %>% filter(name !="Total") %>% ggplot() + geom_line(aes(x=Year,y=value/sum(value),col=name)) +
  labs(title="Worldwide Overdoes Deaths by Drug and Year",y=" % of Deaths Do TO Overdose") + scale_y_continuous(labels=percent) + guides(color = guide_legend(override.aes = list(size = 2)))
```

## US Overdose Deaths from Opioid, Cocaine, Amphetamine and Other Drugs

```{r}
overdose_us <-overdose_data %>%   filter(Country =="United States")
```

# Plot Total Drug Overdose Death by Year

```{r}
p0 <-overdose_us %>% ggplot(aes(x=Year,y=Total)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="USA Total Deaths Do To Drug Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p0)
```
## Percent by Overdose Deaths by Year

```{r}
od_longer %>% filter(name !="Total") %>%
group_by(Year) %>% mutate(Total = sum(value)) %>%
  ggplot() + geom_line(aes(x=Year,y=value/Total,col=name))  +
    labs(title="Worldwide Pecent of OD Deaths by Drug and Year",y="% of Total OD Deaths(log10)")  + guides(color = guide_legend(override.aes = list(size = 2))) + scale_y_log10(labels=percent)
  

```

# USA Drug Overdose Deaths by Drug and Year

```{r}
p1 <-overdose_us %>% ggplot(aes(x=Year,y=Opioid)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="USA Total Deaths Opioid Drug Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p1)
```
```{r}
p2 <-overdose_us %>% ggplot(aes(x=Year,y=Cocaine)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="USA Total Deaths  Drug Cocaine Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p2)
```
```{r}
p3 <- overdose_us %>% ggplot(aes(x=Year,y=Amphetamine)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="USA Total Deaths  Amphetamine Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p3)
```
```{r}
p4 <- overdose_us %>% ggplot(aes(x=Year,y=Other_drug)) + geom_col() +
  scale_y_continuous(labels=comma) +
  labs(title="USA Total Deaths Other Drugs Overdoses",y="Total Drug Overdose Deaths")
ggplotly(p4)
```

## Stacked Plot of OD deaths by Drug

```{r}
us_longer <- overdose_not_world %>% filter(Country =="United States") %>% pivot_longer(              cols=c("Opioid","Cocaine","Amphetamine","Other_drug","Total"))
```

```{r}
us_longer %>% filter(name !="Total") %>% ggplot() + geom_line(aes(x=Year,y=value,col=name)) + scale_y_log10(labels=comma) +
  labs(title="USA overdoes Deaths by Drug and Year",y="Deaths Do TO Overdose") + guides(color = guide_legend(override.aes = list(size = 2)))
```

```{r}
us_longer %>% filter(name !="Total") %>% ggplot() + geom_line(aes(x=Year,y=value/sum(value),col=name)) +
  labs(title="USA Overdoes Deaths by Drug and Year",y=" % of Deaths Do TO Overdose") + scale_y_continuous(labels=percent) + guides(color = guide_legend(override.aes = list(size = 2)))
```

## Percent by Overdose Deaths by Year

```{r}
us_longer %>% filter(name !="Total") %>%
group_by(Year) %>% mutate(Total = sum(value)) %>%
  ggplot() + geom_line(aes(x=Year,y=value/Total,col=name))  +
  scale_y_continuous(labels=percent) +
  labs(title="US Pecent of OD Deaths by Drug and Year",y="Percent of Total OD Deaths")  + guides(color = guide_legend(override.aes = list(size = 2)))
  

```

