---
title: 'OWID: Opioid Deaths Part 1'
author: "David Jackson"
date: "02/01/2021 8:00 PM/EST"
output: 
  html_document: 
    fig_width: 9
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(scales)
library(plotly)
library(dplyr)
library(readr)
library(tidyr)
library(forcats)
library(skimr)
library(pracma) # Various moving averages
```

## Our world In Data (http://ourworldindata.org)
* "Opioid overdoses were by far the most common in the United States, at 13.3 deaths per 100,000. This was twice as high as the second country – Libya – at 7 deaths per 100,000. And the US rate is up to 40-times higher than in some countries in Europe (like Italy, Poland or Hungary)".

```{r}
opioid_rate <- read_csv("./DATA/death-rate-from-opioid-use.csv") %>%
  rename("Deaths" = "Opioid_deaths") %>% 
  rename("Country" = "Entity") 
opioid_rate %>% select(Deaths) %>% skim() # Add to Part 2
```

### Calc and Plot Yearly Death Rate

```{r}
opioid_rate$Total <- opioid_rate$Deaths * 100000

opioid_yearly <- opioid_rate %>% group_by(Year) %>% summarise(
                        Totalper = sum(Deaths),
                        AvgYear = mean(Deaths),
                        TotalYear = sum(Total),
                        TotalAVg = mean(Total)
)
```
```{r}
opioid_yearly %>% ggplot() + geom_line(aes(x=Year,y=Totalper)) +
  scale_y_continuous(labels=comma) + 
  labs(title="Worldwide Opioid Deaths by Year",y=" Deaths per 100,000")
```
```{r}

opioid_yearly %>% ggplot() + geom_line(aes(x=Year,y=TotalYear)) +
  scale_y_continuous(labels=comma) + 
  labs(title="Worldwide Opioid Deaths by Year")

```
```{r}
country_totals <- opioid_rate %>% group_by(Country) %>% summarise(
                        Totalper = sum(Deaths),
                        AvgYear = mean(Deaths),
                        TotalYear = sum(Total),
                        TotalAVg = mean(Total)
)
```

### Calc and plot Top 10 Countries (Death Rate)

```{r}
country_totals %>% filter(Totalper >=87.00965) %>% 
  ggplot() + geom_col(aes(x=reorder(Country,Totalper),y=Totalper)) + coord_flip() +
  scale_y_continuous(labels=comma) + labs(title="Top 10 Countries by Deaths",
                                            y="Deaths Per 100,000")
```
### Numbers by Selected Country

```{r}
opioid_rate %>% filter(Country %in% c("Libya","United States","Russia","North America",
                                      "United Arab Emirates","South Africa","Eastern Europe",
                                      "Estonia","Scotland","Southern Sub-Saharan Africa")) %>%
  
  ggplot() + geom_col(aes(x=Year,y=Deaths,fill=Deaths)) + facet_wrap(~Country,scale="free_y",ncol=2) +
  labs(title="Top 10 Countries by Deaths",y="Deaths per 100,000")
```
### Plots of US Opioid Deaths

```{r}
opioid_rate %>% filter(Country =="United States") %>%
  ggplot() + geom_col(aes(Year,y=Deaths,col=Deaths),lwd=2) + labs(title="Us Opioid Deaths per 100,000")
```
```{r}
opioid_rate %>% filter(Country =="United States") %>%
  ggplot() + geom_col(aes(Year,y=Total,col=Total),lwd=2) + 
  labs(title="Us Total Opioid Deaths by Year(1990 - 2016") +
  scale_y_continuous(labels = comma)
```
